Efficient toolkit implementing best practices for principal component analysis of population genetic data

### 2019-12-13 paper-bedpca.tex
save

\abstract{
Principal Component Analysis (PCA) of genetic data is routinely used to infer ancestry and control for population structure in various genetic analyses. However, conducting PCA analyses can be complicated and has several potential pitfalls. These pitfalls include (1) capturing Linkage Disequilibrium (LD) structure instead of population structure, (2) projected PCs that suffer from shrinkage bias when projecting PCA from a reference dataset to another independent dataset, (3) detecting sample outliers, and (4) uneven population sizes. In this work, we explore these potential issues when using PCA, and present efficient solutions to these. Following applications to the UK Biobank and the 1000 Genomes project datasets, we make recommendations for best practices and provide efficient and user-friendly implementations of the proposed solutions in R packages bigsnpr and bigutilsr.

For example, we show that PC19 to PC40 in the UK Biobank capture LD structure. Using our automatic algorithm for removing long-range LD regions, we recover 16 PCs that capture population structure only.
Therefore, we recommend using only 16-18 PCs from the UK Biobank. 
We provide evidence for a shrinkage bias when projecting PCs computed with data from the 1000 Genomes project.  Although PC1 to PC4 suffer from only moderate shrinkage (1.01-1.09), PC5 (resp.\ PC10) for example suffers from a shrinkage factor of 1.50 (resp.\ 3.14). 
We provide a fast way to project new individuals that is not affected by this shrinkage bias. 
We also show how to use PCA to restrict analyses to individuals of homogeneous ancestry.

@@

Principal Component Analysis (PCA) has been widely used in genetics for many years and in many contexts. For instance, adding PCs as covariates has been routinely used to adjust for population structure in Genome-Wide Association Studies (GWAS) \cite[]{price2006principal,novembre2008interpreting}. PCA has also been used to detect loci under selection \cite[]{galinsky2016fast,luu2017pcadapt}.
Recently, the advent of large population-scale genetic datasets, such as the UK biobank data, has prompted research on developing scalable algorithms to compute PCA on very large data \cite[]{bycroft2017genome}.
It is now possible to efficiently approximate PCA on very large datasets thanks to software such as FastPCA (fast mode of EIGENSOFT), FlashPCA2, PLINK 2.0 (approx mode), bigstatsr/bigsnpr, TeraPCA and ProPCA \cite[]{galinsky2016fast,abraham2017flashpca2,chang2015second,prive2017efficient,bose2019terapca,agrawal2019scalable}.

However, some pitfalls related to PCA of genotype data have been documented and not all of the currently available software address all of these. 
In the following, we outline common pitfalls and explain when they are relevant.
First, some of the PCs may capture LD structure rather than population structure \cite[]{price2008long,abdellaoui2013population,prive2017efficient}. Including PCs that capture LD as covariates in genetic analyses can lead to reduced power, e.g.\ for detecting genetic association within these LD regions in GWAS.  
Second, another issue may arise when projecting PCs of a reference dataset to another study dataset: projected PCs are shrunk towards 0 in the new dataset \cite[]{lee2010convergence,wang2015improved,zhang2019fast}. This shrinkage makes it potentially dangerous to use the projected PCs for analyses such as PC regression, ancestry detection and correction for ancestry.
Third, PC scores may capture outliers that are due to family structure, population structure or other reasons; it might be benificial to detect and remove these individuals to maximise the population structure captured by PCA (in the case of removing a few outliers) or to restrict analyses to genetically homogeneous samples (e.g.\ ``White British'' people in the UK Biobank).
Finally, efficient methods for PCA use approximations, which can results in some lack of precision of computed PCs. This has been shown to be a possible issue for software such as FastPCA and PLINK 2.0, but not for FlashPCA2 and bigstatsr/bigsnpr \cite[]{abraham2017flashpca2,prive2017efficient}.
An overview of existing methods with their respective advantages and limitations is presented in table \ref{tab:method-overview}.

@@

\begingroup
\renewcommand*{\arraystretch}{2}
\begin{table}[h]
\centering
\caption{Overview of existing methods.} 
\label{tab:method-overview}
\begin{tabular}{|c|c|c|c|c|}
\hline
Analysis & Method and/or Software & Citation & Advantages & Limitations \\
\hline
\multirow{3}{*}{PCA} & bigstatsr / bigsnpr & \cite[]{prive2017efficient} & \makecell{Fast and accurate + handle dosages \\ + thinning options directly included} & \makecell{Own format without missing values \\ (fast functions are available \\ for converting and imputing)} \\
 & FlashPCA2 & \cite[]{abraham2017flashpca2} & Fast and accurate & Not parallelised \\
 & PLINK 2.0 (reimplementation of FastPCA) & \cite[]{galinsky2016fast,chang2015second} & Fast & \makecell{Possible lack of accuracy \\ \cite[]{abraham2017flashpca2,prive2017efficient}} \\
\hline
\makecell{Detection of \\ oultier samples} & \makecell{``6 SDs from the mean'' \\ in EIGENSOFT} & \cite[]{patterson2006population} & Simple & Assumes a Gaussian distritbution \\
\hline
\makecell{Detection of \\ homogeneous \\ samples} & R package aberrant & \cite[]{bellenguez2011robust} & Robust & Uses only two statistics at once \\
\hline
\multirow{5}{*}{\makecell{Projection of \\ new individuals \\ onto reference \\ PCA space}} & Simple projection (multiplication by loadings) & & Simple & Shrinkage biased \\
  & Bias-adjusted projection in R package hdpca & \cite[]{dey2019asymptotic} & Independent of new samples & \makecell{Assumes same shrinkage for all individuals \\ + model-based \\ + need all eigenvalues of reference} \\
  & \makecell{Augmentation, Decomposition and Procrustes (ADP) \\ transformation in LASER 2.0} & \cite[]{wang2015improved} & Accurate & Slow (a new PCA for each new sample) \\
  & Online ADP (OADP) in python package FRAPOSA & \cite[]{zhang2019fast} & Much faster than ADP & Does not work for related individuals (Section \ref{proj-related}) \\ 
\hline
\end{tabular}
\end{table}
\endgroup

@@

When applying our automatic procedure to remove long-range LD regions, it does not converge after 5 iterations for the UK Biobank, meaning that it keeps detecting long-range LD regions at each iteration (represented by peaks in PC loadings).
Therefore, we are able to capture only 16 PCs that show stratification that is not LD structure (Figures \ref{fig:UKBB-screeplot}, \ref{fig:UKBB-scores} and \ref{fig:UKBB-loadings}).
Similarly, PC loadings reported by the UK Biobank clearly show that PC19 to PC40 capture LD structure, which is also the case for PC16 and PC18, although less pronounced (see peaks in figure \ref{fig:UKBB-loadings40}).

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.9\textwidth]{UKBB-loadings1-40.pdf}}
\caption{Principal Component (PC) loadings 1 to 40 reported by the UK Biobank.
\label{fig:UKBB-loadings40}}
\end{figure}

@@

To detect a few sample outliers, we compare the standard rule of ``6 SDs from the mean'' (6SD) used in e.g.\ EIGENSOFT to the statistic we propose in section \ref{outlier-sample}.
Our statistic identifies only isolated samples or isolated pairs that seems to be outliers driving structure of PC17-20 of 1000G (Figure \ref{fig:outlier-pd}). 
All but one outliers are distantly-related pairs that disappear if using a more stringent threshold on relatedness (i.e.\ using a KING-relatedness cutoff of \textasciitilde0.0442 instead of \textasciitilde0.0884, see tutorial). 
In contrast, rule 6SD identifies a lot of outliers, of which some are part of a relatively large cluster (Figure \ref{fig:outlier-sd}).
We recall that, in theory, if all PCs are normally distributed, after correcting for multiple testing of 2500 individuals and 20 PCs, 6SD has a small probability of 0.0001 of detecting one outlier or more.

\begin{figure}[htbp]
\centering
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics[width=\textwidth]{hist-outliers-1000G.pdf}
\caption{Distribution of statistics}
\end{subfigure}
\\~\\
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{outliers-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by the statistic (S) used to define outliers.\\~\\}
\end{subfigure}
~~~~~~
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{be-outlier-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier. Threshold of being an outlier is determined based on histogram (a).}
\end{subfigure}
\caption{Outlier detection in the 1000 Genomes (1000G) project, using \texttt{prob\_dist} (Section \ref{outlier-sample}).\label{fig:outlier-pd}}
\end{figure}

@@

\subsection{Projecting PCs from a reference dataset}

@@

As we compute PCA on data with missing values, although we restrict to variants with less than 1\% missing values, we analyse hereinafter the effect of imputation of missing values before computing PCA.
We compare 4 different imputation methods on two different sets of individuals, either all genotyped individuals or all imputed individuals. The UK Biobank imputed data is composed of approximately 1000 individuals less who have been removed because of a high number of missing values or a high heterozygosity \cite[]{bycroft2017genome}.
When computing PCA with mean imputation and using all individuals, PC16 is capturing individuals with very high heterozygosity (Figure \ref{fig:UKBB-het}). When restricting to imputed individuals only, i.e.\ after removing individuals with very high heterozygosity, PC16 completely disappears and new PC16 to PC19 correspond to previous PC17 to PC20 (Table \ref{tab:pca-het}).
When using dosage data instead of genotype data with mean imputation, PCA is globally unchanged (Table \ref{tab:pca-dosage}). 
Overall, if we choose to use either one of the following 4 imputation methods: mean imputation, random imputation according to allele frequencies, using reported dosage data from BGEN files, or imputation of genotyped data based on machine learning using function \texttt{snp\_fastImpute} of R package bigsnpr \cite[]{prive2017efficient}, resulting PCs are always very similar (absolute correlation larger than 0.99 for the 20 computed PCs, results not shown). 
Performing PCA with mean imputation directly on PLINK bed files has the advantage to be fast as compared to imputation using \texttt{snp\_fastImpute}, which took 4 days for 406,545 individuals and 240,444 variants, or based on external reference datasets to get dosages.

@@

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.4\textwidth]{PCA-pipeline.pdf}}
\caption{Pipeline for computing Principal Components (PCs) using R packages bigsnpr and bigutilsr.
\label{fig:pipeline}}
\end{figure}

In this work, we have compiled different pitfalls that can arise with analyses related to Principal Component Analysis (PCA) of genetic data. Then, we have investigated possible solutions to these pitfalls and chose the ones that we found most relevant, both with respect to properties such as accuracy and robustness, but also that would be fast and easy to use. We then implemented these solutions in open-source R packages. 
The new functions we provide can be directly applied to genotypes stored as PLINK bed/bim/fam files with some missing values.
This contrasts with previous releases of package bigsnpr that could only use one custom data format called ``bigSNP''. This data format can store both genotype calls and dosages, but requires conversion from other formats and imputation of missing values using functions provided in the package \cite[]{prive2017efficient}.
As PCA is a useful tool on its own and does not require extensive imputed data, we therefore decided that operating directly on PLINK files with a few missing values would be more practical for users.

@@

We also provide some recommendations when the dataset is composed mainly of one population (e.g.\ British people in the UK Biobank).
We show that it is useful to actually subset these individuals so there is not too much imbalance between the sample sizes of the different populations. Previous work has also shown that uneven population sizes can distort PCs \cite[]{novembre2008interpreting,mcvean2009genealogical}. 
Indeed, when subsetting British and Irish people, we are able to capture a lot more PCs that show population structure with less than 50K individuals as compared to when using more than 400K individuals who are mostly composed of British and Irish people.
This makes designs such as the 1000 Genomes project, which gathered around 100 people for each of 26 populations, highly relevant for capturing population structure.
In contrast, when the goal is to have an homogeneous sample, we show that using the Mahalanobis distance on PC scores can efficiently achieve this goal, which we used in previous analyses \cite[]{prive2019efficient}.
When the homogeneous sample is not predominant in the dataset, one solution is to compute the center and covariance of the robust Mahalanobis distance using only the population of interest, and then computing the distances for all individuals using these robust estimates.

@@

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.9\textwidth]{UKBB-loadings.pdf}}
\caption{Principal Component (PC) loadings 1 to 20 computed on the UK Biobank using \texttt{bed\_autoSVD}.
\label{fig:UKBB-loadings}}
\end{figure}

@@

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.9\textwidth]{projUKBB-related.png}}
\caption{Principal Component (PC) scores 27 to 50 of the UK Biobank.
Black points are the 48,942 individuals of diverse ancestries () used for computing PCA.
These individuals are the ones resulting from removing all related individuals and randomly subsampling the British and Irish individuals.
Red points are the remaining UKBB individuals, projected by simply multiplying their genotypes by the corresponding PC loadings.
Blue points are the remaining UKBB individuals, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation.
Note that only 20,000 random projected individuals are represented in this plot.
\label{fig:projUKBB-related}}
\end{figure}

>>>>

\abstract{	
Principal Component Analysis (PCA) of genetic data is routinely used to infer ancestry and control for population structure in various genetic analyses. However, conducting PCA analyses can be complicated and has several potential pitfalls. These pitfalls include (1) capturing Linkage Disequilibrium (LD) structure instead of population structure, (2) projected PCs that suffer from shrinkage bias, (3) detecting sample outliers, and (4) uneven population sizes. In this work, we explore these potential issues when using PCA, and present efficient solutions to these. Following applications to the UK Biobank and the 1000 Genomes project datasets, we make recommendations for best practices and provide efficient and user-friendly implementations of the proposed solutions in R packages bigsnpr and bigutilsr.

For example, we find that PC19 to PC40 in the UK Biobank capture complex LD structure rather than population structure. Using our automatic algorithm for removing long-range LD regions, we recover 16 PCs that capture population structure only. Therefore, we recommend using only 16-18 PCs from the UK Biobank to account for population structure confounding. We also show how to use PCA to restrict analyses to individuals of homogeneous ancestry.  Finally, when projecting individual genotypes onto the PCA computed from the 1000 Genomes project data, we find a shrinkage bias that becomes large for PC5 and beyond. We then demonstrate how to obtain unbiased projections efficiently using bigsnpr.

@@

Principal Component Analysis (PCA) has been widely used in genetics for many years and in many contexts. For instance, adding PCs as covariates is routinely used to adjust for population structure in Genome-Wide Association Studies (GWAS) \cite[]{price2006principal,novembre2008interpreting}. PCA has also been used to detect loci under selection \cite[]{galinsky2016fast,luu2017pcadapt} and in heritability analyses \cite{yang2010common,loh2015contrasting}.
Recently, the advent of large population-scale genetic datasets, such as the UK biobank data, has prompted research on developing scalable algorithms to compute PCA on very large data \cite[]{bycroft2017genome}.
It is now possible to efficiently approximate PCA on very large datasets thanks to software such as FastPCA (fast mode of EIGENSOFT), FlashPCA2, PLINK 2.0 (approx mode), bigstatsr/bigsnpr, TeraPCA and ProPCA \cite[]{galinsky2016fast,abraham2017flashpca2,chang2015second,prive2017efficient,bose2019terapca,agrawal2019scalable}.

However, in practice, conducting PCA on genotype data to capture population structure consists of more steps than simply performing singular value decomposition on the genotype matrix. These includes removing related individuals, pruning variants in LD, and excluding outlier samples that can suggest poor genotyping quality. Indeed, many pitfalls related to PCA of genotype data have been documented and none of the currently available software address all of these. 
In the following, we outline these pitfalls and explain when they are relevant.
First, some of the PCs may capture LD structure rather than population structure \cite[]{price2008long,abdellaoui2013population,prive2017efficient}.
Including PCs that capture LD as covariates in genetic analyses can lead to reduced power for detecting genetic association within these LD regions.  
Second, another issue may arise when projecting PCs of a reference dataset to another study dataset: projected PCs are shrunk towards 0 in the new dataset \cite[]{lee2010convergence,wang2015improved,zhang2019fast}. This shrinkage makes it potentially dangerous to use the projected PCs for analyses such as PC regression, ancestry detection and correction for ancestry.
This is also an issue when projecting individuals from the same dataset that were discarded from the PCA computation (e.g.\ related individuals).
Third, PC scores may capture outliers that are due to family structure, population structure or other reasons; it might be beneficial to detect and remove these individuals to maximise the population structure captured by PCA (in the case of removing a few outliers) or to restrict analyses to genetically homogeneous samples (e.g.\ ``White British'' people in the UK Biobank).
Finally, efficient methods for PCA use approximations, which can results in some lack of precision of computed PCs. This has been shown to be a possible issue for software such as FastPCA and PLINK 2.0, but not for FlashPCA2 and bigstatsr/bigsnpr \cite[]{abraham2017flashpca2,prive2017efficient}.
An overview of existing methods with their respective advantages and limitations is presented in table \ref{tab:method-overview}.

@@

\begingroup
\renewcommand*{\arraystretch}{2}
\begin{table}[h]
\centering
\caption{Overview of existing methods.} 
\label{tab:method-overview}
\begin{tabular}{|c|c|c|c|c|}
\hline
Analysis & Method and/or Software & Citation & Advantages & Current limitations \\
\hline
\multirow{3}{*}{PCA} & bigstatsr / bigsnpr & \cite[]{prive2017efficient} & \makecell{Fast and accurate + handle dosages \\ + thinning options directly included} & \makecell{Own format without missing values \\ (fast functions are available \\ for converting and imputing)} \\
 & FlashPCA2 & \cite[]{abraham2017flashpca2} & Fast and accurate & Not parallelised \\
 & PLINK 2.0 (reimplementation of FastPCA) & \cite[]{galinsky2016fast,chang2015second} & Fast & \makecell{Possible lack of accuracy \\ \cite[]{abraham2017flashpca2,prive2017efficient}} \\
\hline
\makecell{Detection of \\ oultier samples} & \makecell{``6 SDs from the mean'' \\ in EIGENSOFT} & \cite[]{patterson2006population} & Simple & Assumes a Gaussian distritbution \\
\hline
\makecell{Detection of \\ homogeneous \\ samples} & R package aberrant & \cite[]{bellenguez2011robust} & Robust & Uses only two statistics at once \\
\hline
\multirow{5}{*}{\makecell{Projection of \\ new individuals \\ onto reference \\ PCA space}} & Simple projection (multiplication by loadings) & & Simple & Shrinkage biased \\
  & Bias-adjusted projection in R package hdpca & \cite[]{dey2019asymptotic} & Independent of new samples & \makecell{Assumes same shrinkage for all individuals \\ + model-based \\ + need all eigenvalues of reference} \\
  & \makecell{Augmentation, Decomposition and Procrustes (ADP) \\ transformation in LASER 2.0} & \cite[]{wang2015improved} & Accurate & Slow (a new PCA for each new sample) \\
  & Online ADP (OADP) in python package FRAPOSA & \cite[]{zhang2019fast} & Much faster than ADP & Does not work for related individuals (Section \ref{proj-related}) \\ 
\hline
\end{tabular}
\end{table}
\endgroup

@@

When applying our automatic procedure to remove long-range LD regions, it does not converge after 5 iterations for the UK Biobank, meaning that it keeps detecting long-range LD regions at each iteration (represented by peaks in PC loadings).
Therefore, we are able to capture only 16 PCs that show stratification that is not LD structure (Figures \ref{fig:UKBB-screeplot}, \ref{fig:UKBB-scores} and \ref{fig:UKBB-loadings}).
Similarly, PC loadings reported by the UK Biobank clearly show that PC19 to PC40 capture LD structure, which is also the case for PC16 and PC18, although less pronounced (see peaks in figure \ref{fig:UKBB-loadings40}). These include [TODO: FIND REGION -> CLOSE TO HLA?] in PC19.

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.9\textwidth]{UKBB-loadings1-40.pdf}}
\caption{Principal Component (PC) loadings 1 to 40 reported by the UK Biobank.
We expect population structure to be well distributed along the genome so that we get loadings (weights for each variant for each PC) that are normally distributed around 0 (with small effect sizes).
In contrast, long-range LD structure is basically capturing the variation inside one long-range LD region (so localised in the genome), so that we expect the loadings to be very large in that region only (one peak).
X-axis represent positions of variants in the data, and y-axis the value of loadings.
Points are hex-binned to make plotting of such large number of points manageable.
\label{fig:UKBB-loadings40}}
\end{figure}

@@

To detect a few sample outliers, we compare the standard rule of ``6 SDs from the mean'' (6SD) used in e.g.\ EIGENSOFT to the statistic we propose in section \ref{outlier-sample}.
Our statistic identifies only isolated samples or isolated pairs that seems to be outliers driving structure of PC17-20 of 1000G (Figure \ref{fig:outlier-pd}). 
All but one outliers are distantly-related pairs that disappear if using a more stringent threshold on relatedness (i.e.\ using a KING-relatedness cutoff of \textasciitilde0.0442 instead of \textasciitilde0.0884, see tutorial in section ``code availability''). 
In contrast, rule 6SD identifies a lot of outliers, of which some are part of a relatively large cluster (Figure \ref{fig:outlier-sd}).
We recall that, in theory, if all PCs are normally distributed, after correcting for multiple testing of 2500 individuals and 20 PCs, 6SD has a small probability of 0.0001 of detecting one outlier or more.

\begin{figure}[htbp]
\centering
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics[width=\textwidth]{hist-outliers-1000G.pdf}
\caption{Distribution of statistics}
\end{subfigure}
\\~\\
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{outliers-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by the statistic (S) used to define outliers.
A few points with higher values for this statistic S appear as outliers in PC17-20.}
\end{subfigure}
~~~~~~
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{be-outlier-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier. Threshold of being an outlier is determined based on histogram (a).}
\end{subfigure}
\caption{Outlier detection in the 1000 Genomes (1000G) project, using \texttt{prob\_dist} (Section \ref{outlier-sample}).\label{fig:outlier-pd}}
\end{figure}

@@

\subsection{Projecting onto the PCA space from a reference dataset}

@@

As we compute PCA on data with missing values, although we restrict to variants with less than 1\% missing values, we analyse hereinafter the effect of imputation of missing values before computing PCA.
We compare 4 different imputation methods on two different sets of individuals, either all genotyped individuals or all imputed individuals. The UK Biobank imputed data is composed of approximately 1000 individuals less who have been removed because of a high number of missing values or a high heterozygosity \cite[]{bycroft2017genome}.
When computing PCA with mean imputation and using all individuals, PC16 is capturing individuals with very high heterozygosity (Figure \ref{fig:UKBB-het}). When restricting to imputed individuals only, i.e.\ after removing individuals with very high heterozygosity, PC16 completely disappears and new PC16 to PC19 correspond to previous PC17 to PC20 (Table \ref{tab:pca-het}).
When using dosage data instead of genotype data with mean imputation, PCA is globally unchanged (Table \ref{tab:pca-dosage}). 
Overall, if we choose to use either one of the following 4 imputation methods: mean imputation, random imputation according to allele frequencies, using reported dosage data from BGEN files, or imputation of genotyped data based on machine learning using function \texttt{snp\_fastImpute} of R package bigsnpr \cite[]{prive2017efficient}, resulting PCs are always very similar (absolute correlation larger than 0.99 for the 20 computed PCs, results not shown). 
This justifies performing PCA with mean imputation directly on PLINK bed files with a few missing values; this has the advantage to be much faster than having to impute data using \texttt{snp\_fastImpute}, which took 4 days for 406,545 individuals and 240,444 variants, or based on external reference datasets.

@@

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.4\textwidth]{PCA-pipeline.pdf}}
\caption{Proposed pipeline for computing Principal Components (PCs) using R packages bigsnpr and bigutilsr.
\label{fig:pipeline}}
\end{figure}

In this work, we have compiled different pitfalls that can arise with analyses related to Principal Component Analysis (PCA) of genetic data. Then, we have investigated possible solutions to these pitfalls and chose the ones that we found most relevant, both with respect to properties such as accuracy and robustness, but also that would be fast and easy to use. We then implemented these solutions in open-source R packages. 
The new functions we provide can be directly applied to genotypes stored as PLINK bed/bim/fam files with some missing values.
This contrasts with previous releases of R package bigsnpr that could only use format ``bigSNP''. This data format can store both genotype calls and dosages, but requires conversion from other formats and imputation of missing values using functions provided in the package \cite[]{prive2017efficient}.
As PCA is a useful tool on its own and does not require extensive imputed data, we therefore decided that operating directly on PLINK files with a few missing values would be more practical for users.

@@

We also provide some recommendations when the dataset is composed mainly of one population (e.g.\ British people in the UK Biobank).
We show that it is useful to actually subset these individuals so there is not too much imbalance between the sample sizes of the different populations. Previous work has also shown that uneven population sizes can distort PCs \cite[]{novembre2008interpreting,mcvean2009genealogical}. 
Indeed, when subsetting British and Irish people, we are able to capture a lot more PCs that show population structure with less than 50K individuals as compared to when using more than 400K individuals who are mostly composed of British and Irish people.
This makes designs such as the 1000 Genomes project, which gathered around 100 people for each of 26 populations, highly relevant for capturing population structure \cite[]{10002015global}.
In contrast, when the goal is to have an homogeneous sample, we show that using the Mahalanobis distance on PC scores can efficiently achieve this goal, which we used in previous analyses \cite[]{prive2019efficient}.
When the homogeneous sample is not predominant in the dataset, one solution is to compute the center and covariance of the robust Mahalanobis distance using only the population of interest, and then computing the distances for all individuals using these robust estimates.

[TODO: ADD PART ON MORE WORK NEEDED -> optimal subsetting for populations and effect of including long-range LD regions in e.g.\ GWAS.]

@@

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.9\textwidth]{UKBB-loadings.pdf}}
\caption{Principal Component (PC) loadings 1 to 20 computed on the UK Biobank using \texttt{bed\_autoSVD}.
X-axis represent positions of variants in the data, and y-axis the value of loadings.
Points are hex-binned to make plotting of such large number of points manageable.
\label{fig:UKBB-loadings}}
\end{figure}

@@

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.9\textwidth]{projUKBB-related.png}}
\caption{Principal Component (PC) scores 27 to 50 of the UK Biobank.
Black points are the 48,942 individuals of diverse ancestries used for computing PCA.
These individuals are the ones resulting from removing all related individuals and randomly subsampling the British and Irish individuals.
Red points are the remaining UKBB individuals, projected by simply multiplying their genotypes by the corresponding PC loadings.
Blue points are the remaining UKBB individuals, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation.
Note that only 20,000 random projected individuals are represented in this plot.
\label{fig:projUKBB-related}}
\end{figure}
